Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-11-08 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update Heckman and imputation flow

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, 
      | Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
  plain:     "Imputation of Incomplete Multilevel Data with R"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# environment
set.seed(123)
library(dplyr)
library(ggplot2)
```

@@

## Multilevel data

@@

Table 1: Concepts in multilevel methods

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. Sporadic missingness is visualized in Figure XYZ. 
<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
<!-- TODO: Provide an example for one of the case studies below.] -->
<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

Table 2: Concepts in missing data methods

| **Concept**    | **Details**                                                                |
|----------------|----------------------------------------------------------------------------|
| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
|                | across all data entries                                                    |
| MAR            | Missing At Random, where the probability to be missing depends on observed |
|                | information                                                                |
| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
|                | depends on unrecorded information, making the missingness non-ignorable    |
|                | [@rubi76; @meng94].                                                        |
|                |                              |

@@

@@

## Setup
<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(123)         # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(micemd)       # for case study data and imputation cf. heckman models
library(metamisc)     # for case study data
library(broom.mixed)  # for multilevel estimates
```

TODO: add table with predictor matrix values

- -2 = cluster variable 
- 1 = overall effect
- 3 = overall + group-level effect 
- 4 = individual-level (random) and group-level (fixed) effect

@@

<!-- $$ -->
<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
<!-- $$ -->
```{r}
mod <- popular ~ 1 + sex + (1 | school)
```

The estimated effects in the complete data are presented in Table XYZ. We consider the associations in the full data set to be the true associations.

@@

The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

Evaluate the convergence of the algorithm:
```{r pop-ignored-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```
<!-- [TODO: remove the broom.mixed output, use mitml only] -->

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

@@

```{r pop-print, eval=FALSE}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Evaluate the convergence of the algorithm:
```{r pop-predictor-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

Print the estimates:
```{r, eval=FALSE}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Evaluate the convergence of the algorithm:
```{r pop-multilevel-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

Print the estimates:
```{r, eval=FALSE}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ type + age + motor_score + pupil + ct + (1 | name) 
```

@@

<!-- visualized in Figure \ref{}. -->

```{r forest, echo=FALSE}
# fits <-
#   purrr::map_dfr(
#     unique(impact$name),
#     ~ {
#       glm(
#         mort ~ 1 + age + motor_score + pupil + ct,
#         family = "binomial",
#         data = filter(impact, name == .x)
#       )
#     } %>%
#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
#       cbind(name = as.character(.x), .)
#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
# 
# n <- impact %>% group_by(name) %>% 
#   tally()
# 
# fits %>% left_join(n, by = "name") %>% 
#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
#   theme_classic() +
#   scale_size(guide = 'none')
# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# # TODO: missingness prop inside clusters
# # TODO: make row height proportional to pattern frequency
```

We will use the complete data estimates as comparative truth in this tutorial. The estimated effects in the complete data are presented in Table XYZ. 
```{r impact_truth, include=FALSE, echo=TRUE}
fit <- glmer(mod, family = "binomial", data = impact) # fit the model
tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
```

@@

We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables. We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
```{r impact_corr}
plot_corr(dat, rotate = TRUE) # plot correlations 
```

This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`. Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model. For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used. We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:
```{r}
dat <- select(dat, !hypox)  # remove variable
dat <- mutate(dat, motor_score = as.factor(motor_score))
```

@@

As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:
```{r cca}
fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
```

@@

<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
<!-- 1.	Will the complete-data model include random slopes? -->
<!-- 2.	Will the data contain systematically missing values? -->
<!-- 3.	Will the distribution of the residuals be non-normal? -->
<!-- 4.	Will the error variance differ over clusters? -->
<!-- 5.	Will there be small clusters? -->
<!-- 6.	Will there be a small number of clusters? -->
<!-- 7.	Will the complete-data model have cross-level interactions? -->
<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->

@@

<!-- --- -->
Mutate data to get the right data types for imputation (e.g. integer for clustering variable).

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r impact}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred, rotate = TRUE)

@@

Impute the incomplete data
```{r imp_impact, cache=TRUE}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

Evaluate the convergence of the algorithm:
```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

Analyze the imputed data:
```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(
    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
    family = "binomial"
    )) 
# tidy(pool(fit))
# as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```
The estimated effects after imputation are presented in Table XYZ.

@@

TODO: explain exclusion restriction.

Data are simulated and included in the `micemd` package. We will use the following variables:

- `region` Cluster variable,
- `gender` Gender (0=male, 1=female),
- `age` Age of the patient,
- `height` Height in meters,
- `weight` Weight in kilograms,
- `time`` Response time in minutes (inclusion-restriction variable).

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, 
      | Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# environment
set.seed(123)
library(formatR)
library(dplyr)
library(ggplot2)
```

\newpage

@@

@@

```{r clus0,  echo = FALSE, warning = FALSE,cache = TRUE}
concepts <- data.frame(Concept=c("Sample units","Cluster","Hierarchical data","Level-1","Level-2","Hierarchical model","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
                       Details=c("Units of the population from which measurements are taken in a sample, e.g., students.", "Variable that specify the cluster or agruppation, e.g., Classroom","Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.","Variable that varies within a cluster, eg. Test score","Variable that does not vary within a cluster but between, e.g. teacher experience.","Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,"Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)","Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history","Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).","The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
                                 ""))
library(kableExtra)
concepts%>%
  kable( booktabs = T, escape = F,
         caption = "Concepts in multilevel methods",
         col.names = c("Concept","Details"), align = "l")%>%
  column_spec(c(1), width = "3cm")%>%
  column_spec(c(2), width = "12cm")
```

@@

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. <!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] --> <!-- TODO: Provide an example for one of the case studies below.] --> <!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```

<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

```{r mechanism,  echo = FALSE, warning = FALSE,cache = TRUE}
concepts <- data.frame(Concept=c("MCAR","MAR","MNAR"),
                       Details=c("Missing Completely At Random, where the probability to be missing is equal across all data entries.","Missing At Random, where the probability to be missing depends on observed information","Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]"))
library(kableExtra)
concepts%>%
  kable( booktabs = T, escape = F,
         caption = "Concepts in missing data methods",
         col.names = c("Concept","Details"), align = "l")%>%
  column_spec(c(1), width = "3cm")%>%
  column_spec(c(2), width = "12cm")
```

@@

## Imputation workflow
Below we provide a imputation workflow that can be used in general to impute cluster data.

### Main analysis
When imputing clustered data, our first consideration should be the research question and the type of analysis that researchers intend to perform, assuming that non-incomplete values were present. This initial assessment provides valuable insights into the data's structure (see the level table), relationships among variables, and the causes of missingness.

When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g. if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021).

### Exploration of available data
Then it is required to explore available information across clusters. This can involve evaluating the Intra-class Correlation Coefficient (ICC) to determine if a one-level imputation method is sufficient, particularly if there are no substantial differences between clusters. Additionally, it can be done by visualizing the data using available package as **ggmice**, for instance examining density plots can help us understand the variable type (normal, Poisson, categorical) and its range of plausible values. Exploring relationships between covariates through bivariate plots can be valuable in identifying potential interactions or non-linearities.

In the context of missingness, we should assess patterns of systematic and sporadic missing values and explore the relationships between the missingness of an incomplete variable and observable data. This evaluation allows us to consider the plausibility of Missing at Random (MAR) mechanisms over Missing Completely at Random (MCAR) methods and helps in selecting prediction variables to be included in the imputation model (considering overflux).

### Selecting plausible imputation methods
Before to implement multiple imputation it is requiered to evaluate if simpler imputation methods are suffice: 
- For example, certain incomplete variables may not necessitate stochastic imputation methods like MICE. Instead, they can be effectively addressed through deductive imputation, where incomplete values can be inferred from logic and deterministic relationships between variables. This approach is particularly useful for variables that are functions of others, e.g. a person's BMI from their weight and height, determining values for one-level variables from two-level ones, e.g. in the context of Individual Participant Data (IPD), incomplete information can be extracted from metadata, such as deducing incomplete data about abortion in a country where abortion is illegal, or engaging in cross-temporal or protocol's deduction, like imputing missing test values for deceased patients.

- Under monotonic pattern, the use of regression models
- CC when the proportion of missing is minimal <5%.
- Analysis that are robust to missing data, for instance FIML models such as mixed model do not require imputation models when only the outcome variable is missing.An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML). This method does not require the imputation of any missing values. Whereas MI consists of imputation, analyses and pooling steps, FIML analyses the data in a single step. When the assumptions are met the two approaches should produce equivalent results. [REF] As FIML requires specialised software, not all analyses can be performed with standard software. [REF]

- Add table describing 2l- imputation models. like this https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html

### Imputation process

#### Clustering specification
There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: <https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/>) TODO: replace ref.

#### Selection of variables in the imputation model
- Inclusion of outcome variable [Moons] When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]
- Imputation for interaction term, JAV, Machine lerarning methods (Random forest) and imputation separated by treatment group
- Prediction matrix specfication (table)

#### Convergency problems during imputation
- Check log 
- Problems of convergence on some 2l methods can be solved by use of overflux or quickpred. 
- Consider the use of scaled variables Vanburen. 
- Consider to swap imputation mehtod in problematic variables: The use of non parametrical imputation methods such as pmm. or downgrade to 1l methods.
- Also by the specification of post-processing in some variables can alivianate this. 
- Separated imputation process for each endpoint analysis

### Convergency 
Plots and also the prediction plots.

### Sensibility analysis
About the possibility of assumptions about Missing mechanism.

@@

## Setup

<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->

Install non-CRAN packages if necessary:

```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")

```

@@

Set up the R environment and load the necessary packages:

```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(micemd)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```

<!-- library(broom.mixed)  # for multilevel estimates -->

@@

<!-- $$ -->

<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->

<!-- $$ -->

```{r}
mod <- popular ~ 1 + sex + (1 | school)
```

@@

<!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->

@@

Impute the data, ignoring the cluster structure:

```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

<!-- [TODO: remove the broom.mixed output, use mitml only] -->

Analyze the imputations:

```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

@@

```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Analyze the imputations:

```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

Analyze the imputations:

```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:

```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

@@

The estimated effects in the complete data are visualized in Figure \ref{}.

```{r forest, echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(impact$name),
    ~ {
      glm(
        mort ~ 1 + age + motor_score + pupil + ct,
        family = "binomial",
        data = filter(impact, name == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(name = as.character(.x), .)
  ) %>% filter(term %in% c("ctIII", "ctIV/V"))

n <- impact %>% group_by(name) %>% 
  tally()

fits %>% left_join(n, by = "name") %>% 
  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```

<!-- We will use the following estimates as comparative truth in this tutorial: -->

```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:

```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
```

@@

We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables. We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:

```{r impact_corr}
plot_corr(dat, rotate = TRUE) # plot correlations 
```

This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`. Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model. For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used. We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:

```{r}
dat <- select(dat, !hypox)  # remove variable
```

@@

As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:

```{r cca}
fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
```

@@

<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->

<!-- 1. Will the complete-data model include random slopes? -->

<!-- 2. Will the data contain systematically missing values? -->

<!-- 3. Will the distribution of the residuals be non-normal? -->

<!-- 4. Will the error variance differ over clusters? -->

<!-- 5. Will there be small clusters? -->

<!-- 6. Will there be a small number of clusters? -->

<!-- 7. Will the complete-data model have cross-level interactions? -->

<!-- 8. Will the dataset be very large? (fimd, section 7.3)-->

@@

Mutate data to get the right data types for imputation (e.g. integer for clustering variable).

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:

```{r impact}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred, rotate = TRUE)

@@

Impute the incomplete data

```{r imp_impact, cache=TRUE}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response. We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. We simulate data for 5 clusters so that the true values are known. We use the following variables from the dataset:
